---
title: "red_wine"
author: "karthik"
date: "December 25, 2015"
output: pdf_document
---


```{r,message=FALSE,warning=FALSE,echo=FALSE}
library(ggplot2)
library(memisc)
library(GGally)
library(gridExtra)
library(knitr)

```

#Overview of red wind data set

###Dimention of red wind data set
```{r, echo=FALSE}
#dim
wine<-read.csv("C:\\Users\\karthik\\Desktop\\udacity\\p4\\project4\\wineQualityRed.csv")
dim(wine)
```

##Various variable involed in the data set

```{r, echo=FALSE}
#name
names(wine)
```

## Data Structure 
```{r, echo=FALSE}
#variable types
str(wine)
```

##Dependent Variable's (qualtiy) statistics Summary
```{r,echo=FALSE}
#summary of quality 
table(wine$quality)
summary(wine$quality)
```

##Histogram for Quality of Wine
```{r,echo=FALSE}
ggplot(aes(x=quality),data = wine)+geom_histogram(binwidth = 1,fill='white',color="black")+
  geom_vline(xintercept = mean(wine$quality),linetype=1,color="red")+
  geom_vline(xintercept = median(wine$quality)-IQR(wine$quality),linetype='longdash',color='green')+
  geom_vline(xintercept = median(wine$quality)+IQR(wine$quality),linetype='longdash',color='green')+
  scale_x_continuous(breaks=seq(0,10,1))
```


```{r,echo=FALSE}
#box pot quality 
boxplot(wine$quality,xlab="quality")
```

Both from the histogram and stats summary the 50% of the wine are ranked as 
quality with 5 and 6 out of 10. 

```{r,echo=FALSE,warning=FALSE,message=FALSE}
#multiple histogram
p1<-ggplot(aes(x=volatile.acidity),data=wine)+geom_histogram(binwidth = .05)
p2<-ggplot(aes(x=fixed.acidity),data=wine)+geom_histogram()
p3<-ggplot(aes(x=citric.acid),data=wine)+geom_histogram(binwidth = .02)
p4<-ggplot(aes(x=residual.sugar),data=wine)+geom_histogram()
p5<-ggplot(aes(x=chlorides),data=wine)+geom_histogram()
p6<-ggplot(aes(x=free.sulfur.dioxide),data=wine)+geom_histogram(binwidth = .2)
p7<-ggplot(aes(x=total.sulfur.dioxide),data=wine)+geom_histogram()
p8<-ggplot(aes(x=density),data=wine)+geom_histogram()
p9<-ggplot(aes(x=pH),data=wine)+geom_histogram()
p10<-ggplot(aes(x=sulphates),data=wine)+geom_histogram()
p11<-ggplot(aes(x=alcohol),data=wine)+geom_histogram()+xlab("alcohol%")
p12<-ggplot(aes(x=quality),data=wine)+geom_histogram()
grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,ncol=4)
```

# Initial Findings from Histogram 

#### volatile acidity
  + Most of the wine has a volatile acidity concentration between $0.4 g/dm^{3}$
  and $0.53 g/dm^{3}$. More over the data is normaly distributed. 
  
#### fixed acidity
  + Most of the wine has fixed acidity concentration between $7 g/dm^{3}$ and $9     g/dm^{3}$. 
  + There are some outlies which are spread out to $16 g/dm^{3}$.

#### citric.acid
  + There are alot of ${0}$ which means a lot of wine don't has citric acid. It        also make sense too because citric acid is added as a freshner or flavor to 
  wines     not as a key ingredients in cooking wine.  
  + I deffinitly like to investigate more on this lated as adding flavo make any     changes to the quality of the wine. 

#### residual sugar
```{r, echo=FALSE, message=FALSE}
p1<-ggplot(aes(x=residual.sugar),data = wine)+geom_histogram(binwidth = 1,fill='white',color="black")
p2<-ggplot(aes(x=residual.sugar),data = wine)+geom_histogram(fill='white',color="black")+coord_cartesian(xlim = c(0,3.5))+
  xlab("residual.sugar_without_outliers")
grid.arrange(p1,p2,ncol=2)
```

  + Almost most of the wine has sugar contanr less then $10 g/dm^{3}$. 
  + The lookes alot like long tail but once the outlier is been removed the 
  histogram looks normaly distribued. 
  + Interesing all wine sample has atleast $1 g/dm^{3}$. i will investigate,is 
  there any relations between the sugar contain and quality as i read in an    
  artical apart from sugar from fermentation more sugar are added in wine making
  process so done adding more sugar inceases the wine quality?.
  
#### chlorides:
```{r, echo=FALSE, message=FALSE}
p1<-ggplot(aes(x=chlorides),data = wine)+
  geom_histogram(fill='white',color="black")
p2<-ggplot(aes(x=chlorides),data = wine)+geom_histogram(fill='white',color="black")+
  coord_cartesian(xlim = c(0,.15))+xlab("chlorides_without_outliers")
grid.arrange(p1,p2,ncol=2)
```

  + Chloride is a salt. I don't like my wine to tase salty and others too  i  
  guess that is why the amount of salt contain is less. Almost 75% of sample   
  have less then $0.09 g/dm^{3}$. 
  + once the outliers are removed it very clear that the chloride are spread 
  very well normaly spread across simple.
  
#### free and total sulfur dioxide:
  + Free sulfur dioxide is added to prevent microbial growth and the oxidation 
  of wine the amount of sulfur dioxide contain mean value $15.87 mg/dm^{3}$ with
  some seious outlier striching out to $72 mg/dm^{3}$. 
  + The total sulfur dioxide contain has a mean $72 mg/dm^{3}$ which when 
  compaied to only free sulfur dioxide's mean is much high. i will investigate
  does this high sulfur dioxide affect the quality in laster stage.

#### density
  + Depending on the percent alcohol and sugar content the density of wine is 
  close to water i will investigate how density of the wine affects the quality. 
  + One question does quality increses as the density of  the wine moves close 
  towords the density of the water?.
  
#### pH
  +  pH is the sacle is used to measure liquid is acid or base most of wine are 
  have scale feom ${3}-{4}$ so as our sample. 
  + The question is there trend in the quality of wine as the pH level increase
  or deceases?. 
  
#### Alcohol 
  + Alcohol is one of the importain ingredients thats why all wine bottles carry
  lable with percentage of alcohol contant but how much alcohol contain does a 
  good quality wine has? let me do more anlaysis. 

### Transforming variable to check normality.
```{r, echo= FALSE,warning=FALSE,message=FALSE}
p1<-ggplot(aes(x=citric.acid ),data=wine)+
  geom_histogram()
p2<-ggplot(aes(x=citric.acid ),data=wine)+
  geom_histogram()+scale_x_log10()+xlim(.02,.70)+labs(x="log_citric.acid")
grid.arrange(p1,p2,ncol=2)
```
 
 >* dose wine sample has a lot of citric acid with ${0}$ value ?*
 
 *Yes, the number of `citric.acid` with 0 value is ${132}$. So its not that data
 is missing but acutually the alot of wine doesn't have `citric.acid`.let me 
 remove the 0 and do log transformation so that i can find any distribution from
 the histogram.
  * even after the transfomation there isn't any distibution visiable. i will 
  investiage more with a box plot to check for trend.


```{r, eval=FALSE,echo=FALSE}
table(wine$citric.acid==0)
```

## Corealtion between Variables and Quality 

```{r,echo=FALSE,eval=FALSE}

kable(cor(wine[,2:12],wine$quality))
```


|Positive correation     | Negative correation                 |
|:-----------------:     | :-----------------:                 |
|alcohol =${0.476}$      | volatile acidity = ${-0.390}$       |
|sulphates=${0.251}$     | total sulpure di oxide = ${-0.185}$ |
|citric acid=${0.226}$   | density = ${-0.174}$                |
|fixed acidity=${0.124}$ | chlorides = ${-0.057}$              |
|residual sugar=${0.013}$| pH = ${-0.057}$                     |
| ${---}$                | free sulfur dioxide = ${-0.050}$    |

                    
# Bivariant Plot and Analysis 

## Boxpot of differnt Variable with Quality 
```{r, message=FALSE, echo=FALSE, message=FALSE}
wine$quality_factor<-factor(wine$quality)
b2<-ggplot(aes(x=quality_factor, y=fixed.acidity),data=wine)+geom_boxplot()
b3<-ggplot(aes(x=quality_factor,y=volatile.acidity),data=wine)+geom_boxplot()
b4<-ggplot(aes(x=quality_factor,y=citric.acid),data=wine)+geom_boxplot()
b5<-ggplot(aes(x=quality_factor,y=residual.sugar),
           data=subset(wine,wine$residual.sugar<=4))+geom_boxplot()
b6<-ggplot(aes(x=quality_factor,y=chlorides),data=wine)+geom_boxplot()
b7<-ggplot(aes(x=quality_factor,y=free.sulfur.dioxide),data=wine)+geom_boxplot()
b8<-ggplot(aes(x=quality_factor,y=total.sulfur.dioxide),data=wine)+geom_boxplot()
b9<-ggplot(aes(x=quality_factor,y=density),data=wine)+geom_boxplot()
b10<-ggplot(aes(x=quality_factor,y=pH),data=wine)+geom_boxplot()
b11<-ggplot(aes(x=quality_factor,y=sulphates),data=wine)+geom_boxplot()
b12<-ggplot(aes(x=quality_factor,y=alcohol),data=wine)+geom_boxplot()+ylab("alcohol %")
grid.arrange(b2,b3,b4,b5,b6,b7,b8,b9,b10,b11,b12,ncol=4)
```

* If we follow the tread of the median bar in the box plot the we can find if 
any tread between the quality and the other variable.
  * The variable `alcohol` ,`sulphates` , `citric.acid` all have a positive trend. 
  * The variable `volatile` has negative trend.
  * All other variables doesn't have a good variation to have clear picture of 
  there tread. 
  
#### Investigate more on Alcohol ,Sulphates , Citric.acid, Volatile
To have more clear understanding of quality variable the quality is splited into
3 categories **bad** (score [0-4]), **good** (score [5-6]), **best** (score[7-8])

```{r,message=FALSE, echo=FALSE}
wine$quality_bucket<-cut(wine$quality,breaks=c(0,4,6,8),labels = c("bad","good","best"))
a1<-ggplot(aes(x=quality_bucket,y=volatile.acidity),data=wine)+
  geom_boxplot(fill="red")
a2<-ggplot(aes(x=quality_bucket,y=citric.acid),data=wine)+
  geom_boxplot(fill="green")
a3<-ggplot(aes(x=quality_bucket,y=sulphates),data=wine)+
  geom_boxplot(fill="green")
a4<-ggplot(aes(x=quality_bucket,y=alcohol),data=wine)+geom_boxplot(fill="green")+
  ylab("alcohol %")
grid.arrange(a1,a2,a3,a4,ncol=2)
```

  The trends are more clear and very well labeled for good visualisation. 
  The **red color** of `volatile.acidity` shows it has negative trend while the
  **green color** of `citric.acid`, `sulphates`, `alcohol` shows the positive 
  trend with *wine quality*
  
  The volatile acid, sulphate, alchol, and citric acid have high correlation to 
  wine quality. I like to check how these variable are corelated to each other 
  with polted with a scatter plot.
  
```{r, echo=FALSE, message=FALSE}
ggplot(aes(x=volatile.acidity,y=alcohol),data=wine)+geom_point(aes(alpha=1/10))+
  facet_wrap(~quality_bucket)+ylab("alcohol %")
ggplot(aes(y=volatile.acidity,x=citric.acid),data=wine)+
  geom_point(aes(alpha=1/10))+facet_wrap(~quality_bucket)
ggplot(aes(y=sulphates,x=alcohol),data=wine)+
  geom_point(aes(alpha=1/10),position = "jitter")+
  facet_wrap(~quality_bucket)+ylab("alcohol %")
ggplot(aes(y=sulphates,x=citric.acid),data=wine)+
  geom_point(aes(alpha=1/10),position = "jitter")+
  facet_wrap(~quality_bucket)

```

  The plots revel very little relation between the variables and the scatter 
plot with citric acid and volatile acidity showed a clear negative trand. 
## Building  Models 
```{r,echo=FALSE, message=FALSE}
m1<- lm(I(quality)~I(alcohol),data=wine)
m2<- update(m1,~.  +citric.acid)
m3<- update(m2,~. +sulphates)
m4<- update(m3,~. + volatile.acidity)
m5<- update(m4,~. +fixed.acidity+residual.sugar+chlorides+total.sulfur.dioxide+
              density+pH)
mtable(m1,m2,m3,m4,m5)

#cor.test(wine$quality,wine$alcohol)
```

* Even after adding all the variable i couldn't find a good fix. There are 2 
solution to this problem.
  * Add two are more variable to get a new variable and add it to the model. 
  I'm definitly skeptical in doing so since the full model $R-squared={.359}$ 
  which is a just a 2% high when compaired with a model have `alcohol`, 
  `citric.acid`, `sulphates`, `volatile.acidity`.
  * An other solution is to use a non-leaner model or do somthing call train and
  test your model to have a good prediction and of course a good quality wine.  

## Multi-Varient Plot:

The variable such as `pH`, `density` may not directly corelated with the wine 
quality but they are a good measure of properties of wine. i will investigate 
more on them in coming analysis.

```{r,echo=FALSE, warning=FALSE}
ggplot(aes(x=volatile.acidity,y=alcohol),data=wine)+
  geom_point(aes(color=wine$quality_bucket, alpha=1/10))+xlim(0,0.9)+
  geom_smooth(method = "lm")+ylab("alcohol %")
```

The scatter plot deffinitly separate the `best` Vs `good` wine. The best wine 
are the one with high alcohol with low volatile contain.

```{r, echo=FALSE, warning=FALSE,message=FALSE}
ggplot(aes(x=alcohol,y=density),data=wine)+
  geom_point(aes(color=wine$quality_bucket, alpha=1/10))+
  geom_vline(aes(xintercept = 11),linetype="longdash")+geom_smooth(method = "lm")+
  xlab("alcohol %")
```

There is a trend that most of the best quality wine are 
from the graph it is clear that the wine with more alcohol contant and less 
density falls in the `best` quality and the `good` quality wine has 
comparetively less alcohol contain and dinsity to best quality.

```{r, echo=FALSE, warning=FALSE}
ggplot(aes(y=residual.sugar,x=density),data=wine)+
  geom_point(aes(color=wine$quality_bucket,alpha=1/10))+
  ylim(0,4)+geom_smooth(method = "lm")
```

I would see alot of `good` quality wine are below the line which means a `best` 
quality wine are with high sugar contant with lesser density but the 
visulisation is not as good so let me check just sample with best and good.


```{r, echo=FALSE, message=FALSE, warning=FALSE}
s<-subset(wine,wine$quality>4)
p1<-ggplot(aes(y=residual.sugar,x=density),data=s)+
  geom_point(aes(color=s$quality_factor,alpha=1/10))+
  ylim(0,4)+geom_smooth(method = "lm")+theme(legend.position="bottom")

p2<-ggplot(aes(y=density,x=quality_factor),data=s)+geom_boxplot(aes(fill = quality_bucket))+theme(legend.position="bottom")

grid.arrange(p1,p2,ncol=2)
```

From the scatter plot there is deffinatly a trend as the density increases the
residual sugar level increases this led to a simple question.

  >* Is a good quality wine has more density or lesser density? *
  
The box polt show as `density` decreases the `quality` of the wine increases. 
Thus it very much evident that a `best` quality wine has a property of
*high suger contain with less density*. Finding the optimal range can we done in
a future analysis.
  

```{r, echo=FALSE}
ggplot(aes(x=density,y=pH),data=wine)+facet_wrap(~quality_bucket)+
  geom_point(aes(color=quality_bucket,alpha=1/10))
```

I though there might be some relationship between the density and pH but it seem
there are not much to infer between them. 

```{r, echo=FALSE}
ggplot(aes(y=alcohol,x=volatile.acidity),data=wine)+
  geom_point(aes(color=quality_bucket,alpha=1/10))+geom_vline(xintercept = 0.4, linetype="longdash")+geom_hline(yintercept = 10.5,linetype="longdash")+
  geom_smooth(method = "lm")+ylab("alcohol %")
```

The scatter plot shows most of the best quality of the wine follow on high 
alcohol contain and low volatile acidity.


# Final Plots

##Plot One [box plot for *quality* vs *alcohol*]
```{r echo=FALSE ,message=FALSE, warning=FALSE}
ggplot(aes(x=quality_factor,y=alcohol),data = wine)+
  geom_boxplot(aes(fill=quality_bucket))+ylab("alcohol %")
```

The alcohol% has a high positive correaltion with quality. A box plot with 
quality and alcohol% for different quality bucket will give a clear variation 
of alcohol% for each type of quality.
The box plot show that alcohol % is one of the key ingredient in determinig 
the quality of the wine as the variations is very much seen as **higher** the 
*alcohol* contant the *quality* of the wine fall into `best` quality wine 
category.

## Plot two [relationship between *residual_suger* with *density*]
```{r, echo=FALSE, message=FALSE, warning=FALSE}
s<-subset(wine,wine$quality>4)
p1<-ggplot(aes(y=residual.sugar,x=density),data=s)+
  geom_point(aes(color=s$quality_factor,alpha=1/10))+
  ylim(0,4)+geom_smooth(method = "lm")+theme(legend.position="bottom")

p2<-ggplot(aes(y=density,x=quality_factor),data=s)+geom_boxplot(aes(fill = quality_bucket))+theme(legend.position="bottom")

grid.arrange(p1,p2,ncol=2)
```

The both scatter plot and box plot complement each other. Both plot together 
help in finding the properties of the `best` quality wine as sometime it need
more then a single plot to explain a property of a variable in our case quality.
The box polt show, as `density` decreases the `quality` of the wine increases. 
Thus it very much evident that a `best` quality wine has a property of 
*high suger contain with less density* which is infered from the scatter plot. 
Finding the optimal range can we done in a future analysis.

## Plot three [box plot for *quality* vs *alcohol*]

```{r, echo=FALSE, warning=FALSE,message=FALSE}
ggplot(aes(x=alcohol,y=density),data=wine)+
  geom_point(aes(color=wine$quality_bucket,alpha=1/10))+
  geom_vline(aes(xintercept = 11),linetype="longdash")+
  geom_smooth(method = "lm")+xlab("alcohol %")
```
A scatter plot of Alcohol % vs density had a clear trend  but adding quality 
as color helped to find the key properties of the `best` quality wine. 
There is a trend that most of the best quality wine are 
from the graph it is clear that the wine with more alcohol contant and less
density falls in the `best` quality and the `good` quality wine has 
comparetively less alcohol contain and dinsity to best quality.

# Conclusion

From various plots i was able to summarise the property of  the 
**best quality wine**.  The following table contain the propreties 
*best quality wine*.

|Higher contain          | Lower contain                       |
|:-----------------:     | :-----------------:                 |
|High alcohol contain    | Low volatile acide                  |
|High alcohol contain    | Lesser wine density                 |
|High residual sugar     | lesser wine density                 |

#Reflection
The explorative analysis reveals the key factor affecting the quality of wine 
are alcohol,volatile acidity, sulphates, citric acid but the linear model which
i have used for my prediction is not a good model as its coefficient of 
determination was .44. For the further study i would try differnt modeling 
technic such as random forest modeling might give more accuracy. 